---
title: "Get base layers"
author: "Sean Rohan"
date: "5/27/2020"
vignette:  >
  %\VignetteIndexEntry{Get base layers}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Get base mapping layers

The get_base_layers() function retrieves shapefiles and other map features that are typically used to make plots of RACE survey areas. The function transforms layers to a user-specified projection based on a coordinate reference system. Alternatively, a default coordinate reference system can be assigned by the function, based on the region.

As of 5/27/2020, get_base_layers has support for the SEBS shelf survey area, and SEBS+EBS shelf.

## Retrieve layers with projection automatically selected

This chunk of code retrieves layers for the EBS shelf survey area and automatically assigns a projection (Albers Equal Area with standard latitudes at 55 째N and 60 째N, cetner latitude 57.5 째N, center longitude at 170 째W **...this is relevant below**). Layers are returned in a list.
```{r}
library(akgfmaps)

SEBS <- get_base_layers(select.region = "bs.south", set.crs = "auto")
class(SEBS)
names(SEBS)

```

The list returned by get_base_layers contains:
*akland = Alaska land (class sf).
*survey.area = Survey area boundary (class sf).
*bathymetry = Bathymetric contours. EBS shelf default: 50 m, 100 m, 200 m (class sf).
*place.labels = A data frame with labels and plotting locations (class data.frame).
*graticule = Graticule (class sf).
*crs = The coordinate reference system for the layers (class crs).
*plot.boundary = Suggested plot boundary (class data.frame).
*lon.breaks = Latitude breaks corresponding with graticule.
*lat.breaks = Longitude breaks corresponding with graticule.

## Make a map

Here is a map that uses all of the layers except the plot labels.

```{r fig.width=8, fig.height = 6}

(SEBS_plot <- ggplot() +
  geom_sf(data = SEBS$akland) +
  geom_sf(data = SEBS$bathymetry) +
  geom_sf(data = SEBS$survey.area, fill = NA) +
  geom_sf(data = SEBS$graticule, color = "grey70", alpha = 0.5) +
  coord_sf(xlim = SEBS$plot.boundary$x, 
           ylim = SEBS$plot.boundary$y) +
  scale_x_continuous(name = "Longitude", 
                     breaks = SEBS$lon.breaks) + 
  scale_y_continuous(name = "Latitude", 
                     breaks = SEBS$lat.breaks) + 
  theme_bw())

```

## Adding your own layers

Although the map shows the x-axis and y-axis as latitude and longitude with degrees, coordinates for all of the layers are in **meters**.

Therefore, adding text to the map at 58 N, 165 W using coordinates x = -165, y = 58 results in a label centered 165 m (longitudinally) and 58 m (latitudinally) away from the center of the projection.

```{r fig.width=8, fig.height = 6}
my_label <- data.frame(x = -165, y = 58, label = "Thar be\ndragons")

SEBS_plot + geom_text(data = my_label, 
                      mapping = aes(x = x, 
                                    y = y, 
                                    label = label), 
                      color = "red")

```

The akgfmaps package includes a function to convert data frame coordinates to the projection without changing the data frame type.

```{r fig.width=8, fig.height = 6}
my_label <- data.frame(x = -165, y = 58, label = "Thar be\ndragons") %>% 
  akgfmaps::transform_data_frame_crs(in.crs = "+proj=longlat", out.crs = SEBS$crs)

SEBS_plot + geom_text(data = my_label, 
                      mapping = aes(x = x, 
                                    y = y, 
                                    label = label), 
                      color = "red")

```

## Add labels

Formatted placename labels can be added directly to the plot using akgfmaps::add_map_labels(), although it is suggested to add labels manually to avoid formatting issues.

```{r fig.width=8, fig.height = 6}
(SEBS_plot %>% add_map_labels(region = "bs.south", new.places = SEBS$place.labels))

```