---
title: "Making breaks"
author: "Sean Rohan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making breaks}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Finding nice breaks for plotting

The default Jenks breaks used by \code{make_idw_map()} does not always work well, so the \code{eval_plot_breaks()} function can be used to explore algorithmic alternatives. Here, I explore some options for plotting the 2017 yellowfin sole data.

First, I examine some breaks for different algorithms.
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(akgfmaps)
yfs2017 <- akgfmaps::YFS2017
eval_plot_breaks(CPUE = yfs2017$CPUE_KGHA, n.breaks = 5)
```

Now, let's compare kmeans and jenks using a coarse grid.

```{r message=FALSE, warning=FALSE}
# Create IDW for Jenks
yfs.opt1 <- make_idw_map(x = yfs2017,
             region = "bs.all",
             set.breaks = "jenks",
             in.crs = "+proj=longlat", # Set input coordinate reference system
             out.crs = "EPSG:3338", # Set output coordinate reference system
             grid.cell = c(20000, 20000), # 20x20km grid
             key.title = "yellowfin sole")

yfs.opt2 <- make_idw_map(x = yfs2017,
             region = "bs.all",
             in.crs = "+proj=longlat", # Set input coordinate reference system
               out.crs = "EPSG:3338", # Set output coordinate reference system
               grid.cell = c(20000, 20000),  # 20x20km grid
             set.breaks = "kmeans", 
             key.title = "yellowfin sole")

yfs.opt1$plot
yfs.opt2$plot
```

Now, I'll try some user-specified values based on past survey tech memos. I set the upper break limit (1000) below the maximum CPUE from the data (~1207) because \code{make_idw_map()} adjusts the upper and lower limits of the scale to ensure all data fall within the plotted range.

```{r, message=FALSE, warning=FALSE}
yfs.opt3 <- make_idw_map(x = yfs2017,
             region = "bs.all",
            in.crs = "+proj=longlat", # Set input coordinate reference system
             out.crs = "EPSG:3338", # Set output coordinate reference system
             set.breaks = c(0, 25, 100, 250, 500, 1000),
               grid.cell = c(20000, 20000),  # 20x20km grid
             key.title = "yellowfin sole")
yfs.opt3$plot

```

