---
title: "Make inverse distance weighted maps using akgfmaps"
author: "Sean Rohan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to make IDW maps using akgfmaps}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This vignette demonstrates how to create IDW maps using akgfmaps. If you haven't installed akgfmaps already, follow the installation instructions on the \code{README}. Then, load the akgfmaps package.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(akgfmaps)
```

The package and all of the dependencies should load. If they do not, make sure you are running R >= 3.6, and verify that all of the dependencies listed in the \code{DESCRIPTION} file are installed.

## Documentation

Function documentation is accessible using the standard R \code{?} convention (e.g. documentation for \code{make_idw_map} can be accessed using \code{?make_idw_map}).

```{r eval=FALSE}
?make_idw_map
```

## Make IDW maps

Start by loading and inspecting the 2017 EBS and NBS yellowfin sole CPUE data that is included with the package. 

```{r}
# Load 2017 EBS and NBS yellowfin sole data
yfs2017 <- akgfmaps:::YFS2017
head(yfs2017)
```

The data includes the species common name, year, station id, latitude, longitude, and catch-per-unit effort in kilograms per hectare. At minimum, the data must include latitude, longitude, and catch-per-unit effort data to plot CPUE.

The function for performing the IDW and making a plot is \code{make_idw_map}. The function uses gstat to perform the IDW, then raster, stars, and sf for plotting. As arguments to the \code{make_idw_map} function, Users can specify:

 * The domain for IDW and mapping, using the 'region' argument. Currently, maps can automatically be generated standard EBS shelf survey area (region = "bs.south") and full EBS shelf (region = "bs.all").
 * Set break points for plotting, manually or algorithmically, using the 'set.breaks' argument. This can be a numeric vector of breaks (manual) or a character vector passed to \code{classInt::classIntervals) to algorithmically define breaks.
 * Generate a custom extrpolation domain for the output. This is automatically generated for "bs.south" and "bs.all"
 * The resolution of the extrapolation grid using the 'grid.cell' argument. Extrpolation takes longer for riner resolution grids.
 * Add a key.title (e.g. species name).
 * Define an input coordinate reference system. Retain the default setting if input is in degrees latitude and longitude.
 * Define an output coordinate reference system that will be used for plotting (defaults to Albers Equal Area).
 * Set the number of 'neighbor' points used for IDW. Defaults to 4, more neighbors takes longer to run.
 * Data can be passed to the function using a data.frame or vectors. If using a data.frame, column names need to include CPUE_KGHA, LATITUDE, and LONGITUDE.
 
 See function documentation for default arguments. 
 
 Below, we call the make_idw_map argument, specifying:
  * region = "bs.all" (northern + southern EBS), which has a predefined extrapolation area.
  * \code{grid.cell = c(0.02,0.02)


```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height = 6}
opt1 <- make_idw_map(x = yfs2017, # Pass data as a data frame
             region = "bs.all", # Predefined bs.all area
             set.breaks = "jenks", # Gets Jenks breaks from classint::classIntervals()
             in.crs = "+proj=longlat", # Set input coordinate reference system
             out.crs = "EPSG:3338", # Set output coordinate reference system
             grid.cell = c(20000, 20000), # 20x20km grid
             key.title = "yellowfin sole") # Include yellowfin sole in the legend title

```

The \code{make_idw_map} function returns a list containing object which contain information about the region, number of breaks, and key.title, which are used by other functions to manipulate plots and add labels.

```{r}
opt1$region
opt1$n.breaks
opt1$key.title
```

The extrapolation.grid is a stars object that contains **discrete** CPUE values.

```{r}
class(opt1$extrapolation.grid)
```

The continuous.grid is a stars object that contains **continuous** CPUE values.

```{r}
class(opt1$continuous.grid)
```

The last object in the list is a ggplot object.

```{r, warning = FALSE, fig.width=8, fig.height = 6}
opt1$plot
```

Map layers are included in the package and do not need to be provided separately. Changes to text sizing, color palette, and addition of plot labels is done using change_text_size, change_fill_color, and add_idw


## Changing colors

The \code{change_fill_color} function changes the fill color palette using predefined color palettes. Refer to the function documentation for a list of colors. Below, I show two transformations, one to grey scale and one to red, using pipe operators \code{%>%} and passing the new fill color to the \code{new.scheme} argument. If no argument is provided to new.scheme, a color palette selector tool will open in a separate window \code{colorspace::choose_palette()}. The predefined color palettes are colorblind and grayscale friendly.

```{r message=FALSE, warning=FALSE, fig.width=8, fig.height = 6}
# Grey Note that the WHOLE list that was returned by make_idw_map should be passed to the function.
opt1.grey <- opt1 %>% change_fill_color(new.scheme = "grey", show.plot = FALSE)

# Red
opt1.red <- opt1 %>% change_fill_color(new.scheme = "red", show.plot = FALSE)

opt1.grey$plot
opt1.red$plot

```

The returned list objects retain the same structure as the objects from make_idw_map


## Adding labels

Labels can be added by passing a data frame of labels, locations, and label types to the \code{add_idw_labels} function. Alternatively, pre-defined labels can be added to the plot based on the region. Below, pre-defined labels are added to the plot using \code{geom_text} and \code{geom_shadowtext}. Below, labels and a scale bar are added to the greyscale plot using pre-defined settings based on the plot region.


```{r warning=FALSE, fig.width=8, fig.height = 6}
opt1.grey <- opt1.grey %>% add_map_labels()
opt1.grey$plot
```

Custom labels can also be appended to the object, with sizing and orientation based on defined 'types.' In the example below, a data frame is passed with plotting locations. Internally, the \code{add_idw_plot_labels} function transforms long/lat coordinates in eastings to the Albers Equal Area projection of the map. See function documentation for options to turn on/off spatial transformations.

```{r warning=FALSE, message=FALSE, fig.width=8, fig.height = 6}
# ONLY custom label
opt1.red2 <- opt1.red %>% add_map_labels(new.places = data.frame(type = "islands", region = "bs.all", x = -163, y = 57, lab = "Bristol Bay"), lab.replace = TRUE)
opt1.red2$plot

```

```{r warning=FALSE, message=FALSE, fig.width=8, fig.height = 6}
# Add custom label
opt1.red <- opt1.red %>% add_map_labels(new.places = data.frame(type = "islands", region = "bs.all", x = -163, y = 57, lab = "Bristol Bay"), lab.replace = FALSE)
opt1.red$plot

```

## Create map file

After labels and colors are added to the map, the \code{create_map_file()} function can be used to create publication-quality png files. It automatically rescales axis and text labels for some pre-defined plot sizes. Output file names can be specified using file.prefix, or they will be extracted automatically from the input list. The function passes plotting arguments to \code{png()}. The argument \code{try.change_text_size} indicates whether the function should try to resize text based on a pre-defined scaling factor.

```{r warning = FALSE, fig.width=8, fig.height = 6}
opt1.grey %>%
create_map_file(width = 8, # Passed to png()
                height = 6,  # Passed to png() 
                units = "in",  # Passed to png()
                res = 300,  # Passed to png()
                bg = "transparent", # Passed to png()
                file.path = "./", 
                try.change_text_size = TRUE)
```

## Resizing text

The \code{change_text_size()} function can be used resize plot text prior to creating a map if the desired size for the map is not pre-defined.

```{r warning=FALSE, fig.width=8, fig.height = 6}
opt.grey2 <- opt1.grey %>% change_text_size(size.mult = 0.5)
opt.grey2$plot
```

## Exploring break options

See the breaks vignette for details.
