---
title: "Mapping and spatial analysis using sf, raster, and stars "
author: "Sean Rohan"
date: "3/10/2022"
output: 
  slidy_presentation:
    highlight: tango
    df_print: kable
    fig_width: 5
    fig_height: 5
---

<style type = "text/css">

code.r{
  font-size: 14px;
}

pre{
  font-size: 14px;
}
</style>

## Overview

- Advantages and disadvantages of mapping and spatial analysis in R
- Lots and lots of examples
- Older packages: sp, rgdal
- {sf} package
- {raster} package
- {terra} package
- {stars} package

## Advantages of R for mapping and spatial analysis

- Reproducibility
- Transparency
- Transferability
- Efficiency through automation

## Disadvantages of R for mapping and spatial analysis

- Less intuitive than GIS software (in my opinion)
- Lots of packages and 'ecosystems'
- Pros and cons of packages are ambiguous to new users
- Shifting landscape
- Takes longer to do some things

## Load libraries

```{r setup, message = FALSE, warning = FALSE, include = FALSE}
library(akgfmaps)
library(magrittr)
library(ggplot2)
library(here)
library(sp)
library(rgdal)
```

```{r setup2, message = FALSE, warning = FALSE, include = TRUE, eval = FALSE}
library(akgfmaps)
library(magrittr)
library(ggplot2)
library(here)
library(sp)
library(rgdal)
library(shadowtext)
```

## {rgdal} Reading shapefiles, bindings to spatial libraries
*This package will be retired by the end of 2023*

- Provides bindings to Geospatial Data Abstraction Library (GDAL)
- Provides access to PROJ library

```{r read_rgdal}
ebs_strata_rgdal <- system.file("data", "bs_grid_w_corners.shp", package = "akgfmaps") %>% 
  rgdal::readOGR()
```

## {sp}: Classes and Methods for Spatial Data

- Relies on rgdal for bindings to GDAL and access to PROJ
- Different classes for different data types (e.g. SpatialPolygonDataFrame, SpatialPointsDataFrame)
- Data structure: nested lists
- Not based on ISO standards
- Doesn't play nicely with ggplot (but it has gotten better!)
- Less intuitive (my opinion)

## The {sf} package

- Developed by the same folks as sp
- Provides bindings to GDAL and PROJ
- Data structure: Object classes sf and data.frame
- Data structure: uses ISO standard well-known text strings for geometries and coordinate reference system
- Objects work like a data.frame:  filter/subset, merge/join, aggregate/summarize
- Faster than sp

## {sf}: Reading shapefiles

- For nonraster files (e.g., points, polygons, lines)

```{r read_sf font-size: 8px}
ebs_strata <- system.file("data", "ebs_strata.shp", package = "akgfmaps") %>% 
  sf::st_read()
```

## {sf}: Data structure

```{r sf_structure}
print(ebs_strata)
```

## {sf}: Plotting using base R

```{r sf_plotting_base}
plot(ebs_strata)
```

## {sf}: Plotting using ggplot2

- geom_sf(): ggplot2 geometry for sf
- coord_sf(): coordinate system for sf

```{r sf_plotting_ggplot}
ggplot() +
  geom_sf(data = ebs_strata,
          aes(fill = factor(Stratum))) +
    geom_sf_text(data = ebs_strata,
          aes(label = Stratum))
```

## {sf}: Creating sf objects

Example using EBS shelf yellowfin sole CPUE from 2017

```{r st_as_sf, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
print(head(akgfmaps:::YFS2017))

yfs_df <- akgfmaps:::YFS2017 %>%
  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = "EPSG:4326")

print(yfs_df)
```

Plotted using ggplot2:

```{r st_as_sf2, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
ggplot() +
  geom_sf(data = yfs_df,
          aes(color = CPUE_KGHA)) +
  scale_color_viridis()
```


## {sf}: Filtering/subsetting

Same as filtering/subsetting from data.frame objects.

Subset values where CPUE > 0:

```{r filter, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
sebs_strata <- ebs_strata %>%
  dplyr::filter(SURVEY == "EBS_SHELF")

ggplot() +
  geom_sf(data = sebs_strata,
          aes(fill = factor(Stratum)))
```

## {sf}: Coordinate reference systems (CRS)

- Supports PROJ4 and PROJ6 strings
- PROJ6 as well-known text stings

```{r sf_crs1, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE}
sf::st_crs(yfs_df)
sf::st_crs(sebs_strata)
```

## {sf}: Plotting with mis-matched CRS

ggplot2 does an automatic CRS transformation when there is a mismatch among layers.

```{r}
ggplot() +
  geom_sf(data = sebs_strata, fill = NA, color = "black") +
  geom_sf(data = yfs_df,
          aes(color = CPUE_KGHA)) +
  scale_color_viridis()
```

## {sf}: Spatial operations with mis-matched CRS

{sf} DOES NOT do an automatic CRS transformation for mismatched geodetic coordinate systems. However, operations will work if *geodetic* coordinate systems match but *projected* coordinate systems do not.

Attempt to spatially join points where CPUE >0 to the strata they are in.
```{r}
yfs_cpuegt0_df <- yfs_df %>% dplyr::filter(CPUE_KGHA > 0)

print(try(st_join(sebs_strata , 
                  yfs_cpuegt0_df, 
                  join = st_within), 
          silent = TRUE))
```

## {sf}: Transformation/reprojection

```{r sf_transform}
yfs_cpuegt0_df_3338 <- sf::st_transform(yfs_cpuegt0_df, crs = 3338) 

print(head(yfs_cpuegt0_df_3338))
```

Suggestion: Use EPSG standard coordinate reference systems that don't use WGS84 as a pivot


## {sf}: Spatial join

Same as joining/merging from data.frame objects.

```{r joining}
yfs_cpuegt0_df_3338 <- sf::st_join(yfs_cpuegt0_df_3338,
                    sebs_strata,
                    join = st_within)

ggplot() +
  geom_sf(data = sebs_strata) +
  geom_sf(data = yfs_cpuegt0_df_3338,
          aes(color = factor(Stratum)))
```

## {sf}: Spatial join

Lots of spatial join functions:

- st_contains_properly
- st_contains
- st_covered_by
- st_covers
- st_crosses
- st_disjoint
- st_equals_exact
- st_equals
- st_is_within_distance
- st_nearest_feature
- st_overlaps
- st_touches
- st_within

## {sf}: Non-spatial join

Works just like a data frame. Note: This won't work if both objects are sf.

```{r ns_join, echo=TRUE, warning=FALSE}
npac_grid <- system.file("data", "bs_grid_w_corners.shp", package = "akgfmaps") %>% 
  sf::st_read()

yfs_grid <- npac_grid %>%
  dplyr::inner_join(akgfmaps:::YFS2017, by = "STATIONID")

ggplot() +
  geom_sf(data = yfs_grid,
          aes(fill = CPUE_KGHA),
          color = "black") +
  scale_fill_viridis()

```

## {sf}: Intersection

Clip features to the extent of other features.

Survey grid:
```{r load_npac_grid, echo=TRUE, warning=FALSE}
npac_grid <- system.file("data", "bs_grid_w_corners.shp", package = "akgfmaps") %>% 
  sf::st_read()

ggplot() +
  geom_sf(data = npac_grid)
```

## {sf}: Intersection (cont.)

Clip the extent of npac_grid to the survey area:

```{r intersection2, echo=TRUE, warning=FALSE}
npac_grid <- system.file("data", "bs_grid_w_corners.shp", package = "akgfmaps") %>% 
  sf::st_read()

ebs_grid <- sf::st_intersection(npac_grid, ebs_strata)

ggplot() +
  geom_sf(data = sf::st_intersection(npac_grid, ebs_strata))
```

## {sf}: Distance

Calculate great circle or Euclidean distance.

```{r distance1, echo=TRUE, warning=FALSE}
# Great Circle Distance
yfs_4326 <- akgfmaps:::YFS2017 %>% 
  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = "EPSG:4326")
dist_4326 <- sf::st_distance(yfs_4326)

st_is_longlat(akgfmaps:::YFS2017 %>% 
  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = "EPSG:4326") %>% 
  sf::st_transform(crs = 3338))

yfs_3338 <- akgfmaps:::YFS2017 %>% 
  sf::st_as_sf(coords = c("LONGITUDE", "LATITUDE"), crs = "EPSG:4326") %>% 
  sf::st_transform(crs = 3338)

dist_3338 <- sf::st_distance(yfs_3338)

print(dim(dist_4326))
```

## {sf}: Distance (cont.)

Different distances. Why? 

```{r distance2, echo=TRUE, warning=FALSE}
print(dist_4326[1:3,1:3])
print(dist_3338[1:3,1:3])
```

Because 4326 is a geodetic coordinate system and 3338 is a projected coordinate system. st_distance calculates great circle distance for 4326 and Euclidean distance for 3338.

```{r distance3, echo=TRUE, warning=FALSE}
st_is_longlat(yfs_4326)
st_is_longlat(yfs_3338)
```

## {sf}: Area

Calculate area based on Euclidean or great circle geometry.

```{r sf_area, echo=TRUE, warning=FALSE}
sf::st_area(ebs_strata %>% sf::st_transform(crs = 3338))[1:3]
sf::st_area(ebs_strata %>% sf::st_transform(crs = 4269))[1:3]
sf::st_area(ebs_strata %>% sf::st_transform(crs = 4326))[1:3]
```

## {sf}: Length and perimeter

- Use st_length
- Operates on LINESTRING and MULTILINESTRING well-known text.
- Use the st_cast function to convert a POLYGON to a MULTILINESTRING.

```{r sf_length}
st_length(st_cast(ebs_strata, to = "MULTILINESTRING") %>% sf::st_transform(crs = 4269))[1:3] 
st_length(st_cast(ebs_strata, to = "MULTILINESTRING") %>% sf::st_transform(crs = 3338))[1:3]
```

## {sf}: Centroid

Find the center of geometries.

```{r sf_centroid, echo=TRUE, warning=FALSE}
stratum_center <- sf::st_centroid(ebs_strata)
lab_coords <- sf::st_coordinates(stratum_center) %>%
  as.data.frame() %>%
  dplyr::mutate(Stratum = ebs_strata$Stratum)


ggplot() +
  geom_sf(data = ebs_strata,
          aes(fill = factor(Stratum))) +
  shadowtext::geom_shadowtext(data = lab_coords,
                              aes(x = X, y = Y, label = Stratum),
                              color = "black", bg.color = "white")


```

## {akgfmaps}: Access to shapefiles

- Provides simple access to many common survey shapefiles
- Applies transformations to layers to simplify transformation
- Shapefiles shared among regions when possible; filters to retrieve subsets (when possible).
- Current support: AI, GOA, EBS shelf, NBS

```{r get_base_layers, echo=TRUE, warning=FALSE}
sebs_layers <- akgfmaps::get_base_layers(select.region = "sebs")
names(sebs_layers)
```

## {akgfmaps}: Features

- akgfmaps::get_base_layers(): Retrieve layers and default plotting options for a region
- akgfmaps::generate_layer_guide(): Create a pdf that shows all of the layers in the package
- akgfmaps::generate_region_guide(): Create a pdf that shows all of the layers for a region